---
title: "GIS with `sf`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
---

This is an introduction to using R as a fully functional GIS suite (goodbye ArcGIS!) mostly with the help of the `sf` package. We'll first cover the relevant object classes that we will cover, then move on to using different projections and plotting the resulting maps. Then we'll learn how to convert `data.frame`s to `sf` format and load shapefiles. We'll finish by putting it all together by doing some 'fancy' GIS operations to answer a 'complicated' questions: 'how many lobsters were caught per 100 km^2 in each ~~lobster fishing area (LFA)~~ [NAFO zone](https://www.nafo.int/Data/GIS) in the inshore (arbitrary definition of less than 100 km from shore) and offshore zones'.


```{r load packages, message=FALSE, warning=FALSE}
library(sf)             # package for GIS in R, loading shapefiles, projections, etc
library(sp)             # old package for GIS in R
library(here)           # package to assist in finding files
library(rnaturalearth)  # package for basemaps
library(tidyverse)      # dplyr, ggplot2, etc rolled into one
```

# Spatial Classes

## data.frame

So far, we have been dealing with `data.frame` class objects to represent out spatial data. This is totally fine if all we want to do is plot longitude/latitude coordinates on an x/y grid. 

The `rvdata` from the previous exercise is a great example of spatial data represented in a `data.frame`.


```{r}
source(here("DataSetup.R"))

class(rvdata)
```

However, if we want to calculate distances (m), areas (m^2), or other spatial operations (intersects, overlaps, buffers, etc) or if we want to use a special projection (*e.g.* polar projection) to make our map accurately represent our data, it becomes important to use true spatial class objects using either the `sp` or `sf` packages. These packages are an effective way to bind the relevant spatial metadata (*e.g.* projection) to the data.

## `sp` (Spatial*****DataFrame)

The `sp` package used to be the 'go to' spatial package for GIS type operations in R, but has now been mostly replaced by the `sf` package (see performance comparison [here](https://insileco.github.io/2017/09/12/efficiency-of-spatial-intersects-in-r/). Some support packages don't yet support `sf` or may return an `sp` object by default (*e.g.* `rnaturalearth`) for historical and compatibility reasons.

Here we'll use the `rnaturalearth` package to download a polygon of the North American landmass.

```{r load sp data}
world_sp <- ne_countries()
```

Notice that this is not a `data.frame` format, but rather a `SpatialPolygonsDataFrame`. In this case, the data.frame, the projection, the polygons, etc are stored in 'slots' that can be accessed with the `@` symbol.

```{r sp structure}
class(world_sp)
world_sp@data
world_sp@proj4string
```

The takeaway message here is that you may run into an `sp` object in the wild, so it's important to recognize them. Their class name is usually `Spatial*****DataFrame` where `*****` can be polygons, points, etc depending on the nature of the data. 

While the plotting of `sp` objects is possible using 'base' or `ggplot2` style plotting, this is covered elsewhere, You may want to take steps to avoid (see `sf` section below) or convert `sp` objects into `sf` objects. Which is accomplished using one function:

```{r conver to sf}
world <- st_as_sf(world_sp)
class(world)
```

## `sf` data.frame

The `sf` package is replacing the `sp` package not just because it's new and shiny, but because it's **much faster**, it's mostly compatible with the `tidyverse`, and it's data structures are more intuitive (at least to me)

```{r load sf data}
world <- ne_countries(returnclass = "sf")
class(world)
```

# Projections

Using the `sp` or `sf` packages allows you to nimbly convert your spatial data from one projection/coordinate system/datum to another. For more information on how that is handled in R (mostly `sp` based, but still informative) see this [link](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf).

If you're not sure which projection you would like to use, check out [Projection Wizard](http://projectionwizard.org/#) which can help you interactively choose a projection based on the extent of your map and your priorities (i.e. is your representation of area more important than distance?)

For objects that are already loaded, you can see their projection info:

```{r sf structure}
sp::proj4string(world_sp)
st_crs(world)
```

I will often keep a couple proj4 strings on hand to use throughout a project, usually one for converting raw lat/long data from a `data.frame` and one for plotting maps

```{r projections}
latlong <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0" # for converting lats/longs
maritimes_eqdc_proj <- "+proj=eqdc +lat_1=42.0 +lat_2=47.0 +lon_0=-62.0"     # for nice plots in the maritimes
arctic_laea_proj <- "+proj=laea +lat_0=90.0 +lon_0=-100.0"                   # for nice Arctic plots
```

But before we start transforming our spatial data into new projections, let's talk about plotting

# Intro to Plotting

It is totally possible to plot `sf` objects in 'base' plot, but I won't cover this in depth here. I'll just point out that you need to be aware of your projection since you may accidentally plot things where they don't belong!

```{r plot world}
plot(world$geometry, col="burlywood")
```

I personally like the bells and whistles that come with the `ggplot` system.

```{r ggplot world}
ggplot(world) + 
  geom_sf(fill="burlywood",alpha=0.5)
```

We can transform our data ahead of plotting (recommended) or set the projection in `ggplot` with `coord_sf()` but then you need to transform the data every time you plot.


```{r plot Arctic}
Arctic <- world %>% 
  filter(continent %in% c("North America","Asia","Europe")) %>% # dplyr action to select continents
  st_transform(arctic_laea_proj)                                # transform!

ggplot(Arctic ) + 
  geom_sf(fill="burlywood",alpha=0.7)
```

```{r plot Arctic maritimes proj}
ggplot(Arctic) + 
  geom_sf(fill="burlywood",alpha=0.7)+
  coord_sf(crs = maritimes_eqdc_proj)
```

There's a lot more plotting below, but first let's get some data ready so we can try to answer our question.

# Converting a `data.frame` to `sf`

You can convert a `data.frame` to an `sf` object, but you need to tell it which columns your coordinates are in and in what coordinate reference system those are in. Here we assume that they are "`r latlong`"" which is a proj4 string that means we have degrees latitude/longitude and we're using WGS84 as our datum. If this is wrong, it could mean everything could be offset by a few meters or worse, see this [link](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf) for more information.

```{r convert to sf}
sfdata <- plotdata %>% 
  st_as_sf(coords = c("long","lat"),crs=latlong) %>% 
  st_transform(maritimes_eqdc_proj)

lobster <- sfdata %>% 
  filter(species == "AMERICAN LOBSTER")
```


# Reading shapefiles

Using the `sf` package, reading a shapefile is a single command!

```{r read NAFO}
NAFO <- st_read(here("Divisions/Divisions.shp"))
```

But let's filter out unnamed (i.e. `NA`) NAFO zones and transform it to the `maritimes_eqdc_proj` that we'll be using for everything below.

```{r read clean NAFO}
NAFO <- st_read(here("Divisions/Divisions.shp")) %>% 
  filter(!is.na(ZONE)) %>% 
  st_transform(maritimes_eqdc_proj)
```

# GIS operations

## Cropping

Now to answer our lobster question, we don't want to plot the whole world, so let's crop down our shapefiles to make plotting easier and faster.

```{r crop}
Maritimes <- ne_states(country = c("Canada", "United States of America"),returnclass = "sf") %>% 
  st_transform(maritimes_eqdc_proj) %>% # transorm
  st_crop(st_bbox(st_buffer(lobster,10^5))) # crop a bounding box with a 100km buffer around the lobster data

ggplot(Maritimes) + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7)
```


The plot auto-zooms to your data, but this looks a little funny, we can fix that with `coord_sf`. You could actually set `xlim` and `ylim` (being careful about coordinate systems) in `coord_sf`, but in Windows, that can make the fill in your polygons look weird until you save the plot as a `.png` using `type = "cairo"`.

```{r plot with expand}
ggplot(Maritimes) + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7) +
  coord_sf(expand=FALSE)
```

You can add layers of data, just like in ggplot OR your favorite GIS interface.

```{r plot add lobster}
ggplot(Maritimes) + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7) +
  geom_sf(data = lobster) +
  coord_sf(expand=FALSE) 
```

```{r plot add NAFO}
ggplot(Maritimes) + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7) +
  geom_sf(data = NAFO, aes(fill=ZONE)) +
  geom_sf(data = lobster) +
  coord_sf(expand=FALSE) 
```


Clearly, `NAFO` has not been cropped yet, but to calculate area for our lobster question, we need the 'full' areas, not the cropped area.

```{r crop NAFO}
 NAFOcrop <- NAFO %>% 
  st_transform(maritimes_eqdc_proj) %>%        # transform
  mutate(area = st_area(geometry)) %>%         # calculate area
  st_crop(st_bbox(st_buffer(lobster,10^5)))  # crop with 100km buffer around lobster data

ggplot(Maritimes) +
  geom_sf(data = NAFOcrop, aes(fill=ZONE),colour="transparent") + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7)+
  geom_sf(data = lobster) +
  coord_sf(expand=FALSE) 

```

## Spatial operations

The `sf` package has a lot of operations you can do on spatial data like `st_intersects()`, `st_overlaps()`, and `st_covers()` (see their combined [help file](https://www.rdocumentation.org/packages/sf/versions/0.8-0/topics/geos_binary_pred) for more options). These functions return a sparse index list if the shapes meet the logical condition set out by the function (*e.g.* polygon/point/line A overlaps B, or C intersects D).

```{r covers}
st_covers(NAFOcrop,lobster)
```

By itself, this is not yet that useful, but with a little help from `lengths()` we can get a count of lobsters in each NAFO zone which we can then normalize by area.

```{r lobsters per NAFO}
NAFOcrop <- NAFOcrop %>% 
  mutate(lobster1000sqkm=lengths(st_covers(.,lobster))/as.numeric(area)*10^9)


ggplot(Maritimes) +
  geom_sf(data = NAFOcrop, aes(fill=lobster1000sqkm),colour="transparent") + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7)+
  coord_sf(expand=FALSE)
```

## Buffer

See I told you we'd talk about st_buffer again! It creates a buffer around the data. We also use `st_union()` to unify all the `Maritimes` polygons to create a single buffer polygon as opposed to one for each province.

```{r buffer}
buffer <- st_union(Maritimes) %>% 
  st_buffer(10^5)

ggplot(buffer)+geom_sf()
```

But that still includes the 'land' parts, so lets extract the buffer where it intersects with the NAFO zones.

## Intersection

Where 2 shapes exist over one another is called the intersection. We can use `st_intersection()` to extract the **polygons** where the 2 intersect, as opposed to `st_intersect` which would just return the **sparse index list** of where the 2 intersect.

```{r intersection}
buffer <- st_union(Maritimes) %>% 
  st_buffer(10^5) %>% 
  st_intersection(st_transform(NAFO,maritimes_eqdc_proj),.) %>%
  st_sf() %>% 
  mutate(shore="nearshore")

ggplot(buffer)+geom_sf()
```

Now we've got our 'near-shore polygons', but what about the offshore polygons? We can use `st_difference()` which I like to think of as the opposite of `st_intersection()`

```{r difference}
NAFOcropbuffer <- NAFO %>%
  filter(ZONE %in% NAFOcrop$ZONE) %>% 
  st_difference(st_union(buffer)) %>% 
  mutate(shore="offshore")

ggplot(NAFOcropbuffer)+geom_sf()
```

Now let's combine both the the near- and offshore.

```{r combined}
NAFOcropbuffer <- NAFO %>%
  filter(ZONE %in% NAFOcrop$ZONE) %>% 
  st_difference(st_union(buffer)) %>% 
  mutate(shore="offshore") %>% 
  rbind(buffer)

ggplot(NAFOcropbuffer)+
  geom_sf(aes(fill=paste(ZONE,shore)))
```

Finally we are ready to answer our question: 'how many lobsters were caught per 100 km^2 in each ~~lobster fishing area (LFA)~~ [NAFO zone](https://www.nafo.int/Data/GIS) in the inshore (arbitrary definition of less than 100 km from shore) and offshore zones'

```{r final}
NAFOcropbuffer <- NAFOcropbuffer%>% 
  mutate(area = st_area(geometry),
         lobster1000sqkm=lengths(st_covers(.,lobster))/as.numeric(area)*10^9) %>% 
  st_crop(st_bbox(st_buffer(lobster,10^5)))


ggplot(Maritimes) +
  geom_sf(data = NAFOcropbuffer, aes(fill=lobster1000sqkm),colour="transparent") + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7)+
  coord_sf(expand=FALSE)+
  scale_fill_distiller(palette="YlOrRd",direction=1)+
  labs(fill = "Lobsters Per 1000 km^2")
```



This may seem like it took a LOT of coding to get here, but a good portion was demonstrative code that I used for educational purposes. Here's how I would normally code it:

```{r summary, message=FALSE, warning=FALSE}
library(sf)             # package for GIS in R, loading shapefiles, projections, etc
library(sp)             # old package for GIS in R
library(here)           # package to assist in finding files
library(rnaturalearth)  # package for basemaps
library(tidyverse)      # dplyr, ggplot2, etc rolled into one


# set projections -------------------------------------
latlong <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0" 
maritimes_eqdc_proj <- "+proj=eqdc +lat_1=42.0 +lat_2=47.0 +lon_0=-62.0"


# load files -----------------------------

source(here("DataSetup.R"))

lobster <- plotdata %>% 
  st_as_sf(coords = c("long","lat"),crs=latlong) %>% 
  st_transform(maritimes_eqdc_proj) %>% 
  filter(species == "AMERICAN LOBSTER")

Maritimes <- ne_states(country = c("Canada", "United States of America"),returnclass = "sf") %>% 
  st_transform(maritimes_eqdc_proj) %>% # transorm
  st_crop(st_bbox(st_buffer(lobster,10^5)))

NAFO <- st_read(here("Divisions/Divisions.shp")) %>% 
  filter(!is.na(ZONE)) %>% 
  st_transform(maritimes_eqdc_proj)


# creat nearshore vs offshore ----------------------------

buffer <- st_union(Maritimes) %>% 
  st_buffer(10^5) %>% 
  st_intersection(st_transform(NAFO,maritimes_eqdc_proj),.) %>%
  st_sf() %>% 
  mutate(shore="nearshore")

NAFOcropbuffer <- NAFO %>%
  st_difference(st_union(buffer)) %>% 
  mutate(shore="offshore") %>% 
  rbind(buffer) %>% 
  mutate(area = st_area(geometry),
         lobster1000sqkm=lengths(st_covers(.,lobster))/as.numeric(area)*10^9)%>% 
  st_crop(st_bbox(st_buffer(lobster,10^5)))


# plot ----------------------------------

ggplot(Maritimes) +
  geom_sf(data = NAFOcropbuffer, aes(fill=lobster1000sqkm),colour="transparent") + 
  geom_sf(fill="burlywood",colour="burlywood",alpha=0.7)+
  coord_sf(expand=FALSE)+
  scale_fill_distiller(palette="YlOrRd",direction=1)+
  labs(fill = "Lobsters Per 1000 km^2")
```





