---
title: "04 Reading/Plotting/Analysis"
output: html_notebook
---

# Creating a data file

We do this only so that we actually have a file to read

```{r}
data(iris)
write.csv(file='iris.csv', row.names = FALSE)
```


# Reading data

Read the csv file:

```{r}
d <- read.csv('iris.csv')
str(d)
head(d)
```

However, we can also read Excel files (xls and xlsx) using the `readxl` package:

```{r}
library(readxl)
head(read_xlsx('iris.xlsx'))
```
Note than when using `read_xlsx` we get a slightly different kind of data frame -- a "tibble".

The example here is just to see about working with the data, so I'm not going to focus on the xlsx details.

# Working with a data frame

Think of the data frame as the R version of a spreadsheet. We can quickly do some summaries, especially by using the `plot()` command:

```{r}
plot(d)
```

By default, plotting a data frame will plot each column against every other column, in a grid. This can be good for getting a good first look at possible relationships, data quality, etc. For example, it looks like in the `iris` dataset that there is a relationship between petal length and petal width.

## Factors

Factors are a special kind of data type for "categorical" data. E.g. in our `iris` example the `read.csv()` function decided that the `Species` column was probably a "factor". Basically, R encodes the column with numeric "identifiers", while keeping track of what each of the identifiers means. This allows us to make summaries based on those factors, such as a boxplot of sepal width separated by species:

```{r}
boxplot(Sepal.Width ~ Species, data=d)
```

We can be more specific about what to plot by specifying columns,, e.g.:

```{r}
plot(d$Petal.Length, d$Petal.Width)
```

Or, with slightly different syntax (similar to `boxplot()`):
```{r}
plot(Petal.Width ~ Petal.Length, data=d)
```

Or, even use the `with()` function:
```{r}
with(d, plot(Petal.Length, Petal.Width))
```

# Adding columns

Much like in a spreadsheet, we can add columns to our data frame:

```{r}
d$Petal.Area <- d$Petal.Length * d$Petal.Width
boxplot(Petal.Area ~ Species, data=d)
```

# Doing analysis (linear models)

Real analysis comes from working with the data and applying statistical methods. A common task is to fit a linear model (e.g. least squares) to data to quantify relationships. We can do that with the very powerful `lm()` function.

The `lm()` function uses a similar "formula" notation to specify independant and dependent variables to the `boxplot()` and `plot()` functions. We can get a summary of the fit with the generic `summary()` function:

```{r}
model <- lm(Petal.Width ~ Petal.Length, data=d)
summary(model)
```

We can also plot the `model` object:
```{r}
par(mfrow=c(2, 2))
plot(model)
```

And finally, we can make a plot of our data, and add the relationship to it. We can even use the `model` object to make predictions for values that we have no data for, using the `predict()` generic function:

```{r}
plot(Petal.Width ~ Petal.Length, data=d)
abline(model, col='red')
length <- c(2.2, 2.75)
width <- predict(model, newdata=list(Petal.Length=length))
points(length, width, pch=2, cex=2, col='green', lwd=3)
```

