---
title: "The basics of plotting in R"
author: "Clark Richards"
date: "December 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A fundamental part of using any interpreted programming language for data analysis (such as R) is the ability to make plots of the data and the results of the analysis. The standard installation of R contains the ability to make a wide array of plots (through the included `graphics` package), and many other packages exist that extend and improve plotting for certain kinds of analysis or data types.

This workshop will go over the fundamentals of making plots in R, using the "base graphics" system. It will not take long working in the R ecosystem to discover the popularity of the `ggplot2` package, which is a powerful environment for making statistical plots using the paradigm of the "Grammar of Graphics" (see http://vita.had.co.nz/papers/layered-grammar.pdf).

I will not touch on ggplot2 at this point in the series, though depending on the type of analysis and the form of the data, I do often recommend that R users learn it.

# The `plot()` function

Though the `plot()` function seems on the surface like one of the most straightforward functions available in R, the nature of making plots is that there are often a lot of things that a given user would like to customize. Further, the `plot()` function in R is actually a *generic* function -- meaning that it can be programmed to behave differently depending on the class of the object being passed to it. We'll get to that more later on, but we already saw one example last time when we passed a data frame to the `plot()` function -- we get a plot of each column against every other column:
```{r}
data(iris)
plot(iris)
```

## Simple plots

The most straightforward way to make a plot is to use it in the form
```{r eval=FALSE}
plot(x, y)
```
where `x` are the *independent* variables and `y` are the *dependent* variables. The "formula" notation will also work (which is consistent with other R functions, like for fitting models), and
```{r eval=FALSE}
plot(y ~ x)
```
will produce the same result. For some examples, we can use the built-in dataset `iris` (but explicitly pass certain columns), or just make up some of our own data.
```{r}
x <- seq(0, 1, by=0.01)
y <- x^2 + rnorm(x, sd=0.1)
plot(x, y)
```

By default, the `type=` argument is set to `"p"`, which stands for "points". Other options are:

* "p" for points,
* "l" for lines,
* "b" for both,
* "c" for the lines part alone of "b",
* "o" for both 'overplotted',
* "h" for 'histogram' like (or 'high-density') vertical lines,
* "s" for stair steps,
* "S" for other steps, see 'Details' below,
* "n" for no plotting.

To plot with lines do:
```{r}
plot(x, y, type='l')
```

We can change some things about the plot using some of the most-used arguments, including x/y labels (`xlab=`/`ylab=`) and the plotting character (`pch`).
```{r}
plot(x, y, xlab='Snoots', ylab='Snogs', pch=2)
```

For documentation of the different `pch` that are available, see `?pch`. Basically, they are identified by the integers between 1 and 25, and we can view them by making a simple plot:
```{r}
plot(1:25, pch=1:25)
```

Want a title? Easy -- you can either use the `main=` argument in plot, or the separate `title()` function to add it after:
```{r}
plot(x, y, xlab='Snoots', ylab='Snogs', pch=2, main='Snoots and Snogs')
## title('Snoots and Snogs')
```

If you're plotting lines, and you want to change aspects of the line, you can do as:
```{r}
plot(x, y, type='l',
     lwd=3, #line width
     lty=2, #line type
     col='red' #color, obviously
)
```

For a complete (and slightly overwhelming) overview of all the `par`ameters that you can change in plots, do `?par`. If you call `par` with no arguments, it returns a list of all the currently set graphics parameters:
```{r}
str(par())
```


### Adding to plots

You'll notice that calling the `plot()` function always creates a *new* plot -- if you want to *add* things to a plot, you should use the `points()` and `lines()` functions:
```{r}
plot(x, y, type='l')
points(x, y, pch=21, col=2, bg='grey', lwd=2)
lines(x, y + 0.25, lty=2)
lines(x, y + 0.5, lty=3)
```

# Multipanel plots

To make plots with more than one panel (like the default plot for data frames), we can use the `par()` function to set up plot parameters *before* calling the `plot()` function. The parameter we want for this is `mfrow`, which indicates we want a multi-figure plot, filled *row-wise*. An equivalent option is `mfcol`, which fills the panels *column-wise*. To make a plot with 2 rows and 3 columns, we first set the `par` with `par(mfrow=c(2, 3))`. Then each new `plot()` command will fill the next panel:
```{r}
par(mfrow=c(2, 3))
plot(x, y)
plot(x, y, pch=2, col=2)
plot(x, y, pch=3, col=3)
plot(x, y, pch=4, col=4)
plot(x, y, pch=5, col=5)
plot(x, y, pch=6, col=6)
```

Note that the above example would be a nice demonstration of a `for` loop in R:
```{r}
par(mfrow=c(2, 3))
for (i in 1:6) {
    plot(x, y, pch=i, col=i)
}
```

# Adjusting margins

R makes pretty nice looking plots by default -- but one criticism that is common is that there tends to be a lot of whitespace. For single plots this isn't a big deal, but when combining subplots it produces a lot of waste.

To adjust the margins, we again use the `par()` function, but now focus on the two parameters: `mar` and `mgp`.

The `mar` parameter is a vector that gives the margin widths around the plot (in units of *lines*), like: `c(bottom, lhs, top, rhs)`.

The `mgp` parameter is a 3 element vector that describes where margin "stuff" should go -- the 3 elements correspond to the line number for:

1. axis labels,

2. tick labels, and

3. ticks

Typical default values for `mar` and `mgp` are:
```{r}
par('mar')
par('mgp')
```

Let's use the 6 panel plot from above, but try and tighten things up. First I'll change only the `mar`:
```{r}
par(mfrow=c(2, 3),
    mar=c(3, 3, 0.5, 0.5))
for (i in 1:6) {
    plot(x, y, pch=i, col=i)
}
```

That looks better, but notice how the axis labels are gone! We can fix that with `mgp`:
```{r}
par(mfrow=c(2, 3),
    mar=c(3, 3, 0.5, 0.5),
    mgp=c(2, 0.5, 0))
for (i in 1:6) {
    plot(x, y, pch=i, col=i)
}
```

# Plot legends

Legends can be added using the `legend()` function. Read the documentation -- there are quite a few options, but mostly it just takes some practice to get the hang of it. Because of R's "static" graphics system (i.e once something is drawn you can go back and edit that thing), the legend creation is more or less manual -- unlike, say, Matlab, which tries to be smart about what you've plotted and makes the legend auto-magically to correspond to the things you actually did. That's great when it works, but can be very inflexible if you want something a little different. On the other hand, in R it would be very easy to make a legend that is completely meaningless, so some care is warranted.

Easiest is probably to illustrate the use of legend with some examples. Basically, make the plot, and then add the elements (and their particulars) to the legend arguments to identify the different features:
```{r}
plot(x, y, pch=19, col=2)
points(x, y + 0.5, pch=2)
lines(x, (y + y+0.5)/2, pch=3, col=3)
grid() # add grid lines
legend('bottomright',
       legend=c('Original y',
                'y + 0.5',
                'Mean of y and y+0.5'),
       pch=c(19, 2, NA),  # Note use of NA for the "line" feature
       lty=c(NA, NA, 1),  # Note use of NA for the "points" features
       col=c(2, 1, 3),
       title='My awesome legend')
```


# Other useful plot types

* **Histograms:** Histograms are useful in statistical analyses, and can be done using the `hist()` function:
```{r}
len <- c(10, 100, 1000, 10000)
par(mfrow=c(1, length(len)))
for (l in len) {
    hist(rnorm(l), main=paste("N=", l))
}
```


* **Boxplots**: boxplots are another common statistical summary plot, which can be made with the `boxplot()` function. Boxplots are optimized for tabular data (e.g. data frames), and so generally behave sensibly with a minimum of arguments. For example, just doing a `boxplot()` of a data frame gives a summary for each of the columns:
```{r}
boxplot(iris)
```

But we can also use some of the columns (remember how the `Species` column is a "factor"?) to plot different summaries:
```{r}
boxplot(Petal.Width ~ Species, data=iris)
```
