---
title: "Introduction to the grammar of data manipulation using dplyr"
output: html_notebook
---

This code is to introduce you to the use of *dplyr* as a tool for data manipulation. *dplyr* can be broken into verbs which are functions that are interoperable by using 'pipes' (%>%) from the package *magrittr*, *dplyr* is becomes a sequential language of data manipulation that is often much more efficient (computing time and code) than base *R*  

load libraries
```{r}
library(dplyr)
```

load data from a zipped file in the 'data' folder that is part of the bioRworkshops repository
```{r}
rvdata <- read.csv(unz("data/RV2010_2018.zip", "RV2010_2018.csv"), stringsAsFactors = F)
```

source a function for making maps
```{r}
source("12_R_Mapping/makemap.R")
```
##glimpse -------

Inspect the data. *glimpse* is similar to *str* but with a nicer format
```{r}
glimpse(rvdata)
```

##select --------

Lets look at specific columns using the *select* function
```{r}
head(select(rvdata,SETNO,YEAR,LATITUDE,LONGITUDE,TOTWGT,COMM))
```

dplyr has some handy functions for quickly parsing out column names
```{r}
head(select(rvdata,starts_with("XT")))
```

```{r}
head(select(rvdata,ends_with("UDE")))
```

```{r}
head(select(rvdata,contains("TEMP")))
```

Bring it all together as a string
```{r}
head(select(rvdata,starts_with("XT"),ends_with("UDE"),contains("TEMP"),YEAR))
```

##arrange -------

Descending by weight

```{r}
head(arrange(rvdata,-TOTWGT))
```

by year
```{r}
head(arrange(rvdata,YEAR))
```

by year, species, and descending weight
```{r}
head(arrange(rvdata,YEAR,COMM,-TOTWGT),25)
```

##rename -----
rename columns

```{r}
head(rvdata%>%rename(species=COMM,lat=LATITUDE,long=LONGITUDE))
```

## pipes --------

The pipe operator: %>% is imported from another package (magrittr). This operator allows you to pipe the output from one function to the input of another function. Instead of nesting functions (reading from the inside to the outside), 
the idea of of piping is to read the functions from left to right.

R code is inherently nested. Nested code can get messy
```{r}
head(arrange(rvdata,YEAR,COMM,-TOTWGT))
```

but we can streamline with the pipes
```{r}
rvdata%>%arrange(YEAR,COMM,-TOTWGT)%>%head
```

```{r}
rvdata%>%
  select(ends_with("UDE"),contains("TEMP"),YEAR,COMM,TOTWGT)%>%
  arrange(YEAR,COMM,-TOTWGT)%>%
  head
```

remember that the order matters. in this case the code fails because we select the columns of interst before we sort by one we don't keep. 
```{r}
rvdata%>%
  select(ends_with("UDE"),contains("TEMP"),YEAR,COMM,TOTWGT)%>%
  arrange(-DMIN)%>%
  head ## ORDER MATTERS
```

this is fixed with the order of operations. In this case code is transferred sequentially to the right (down) based on the %>% pipes
```{r}
rvdata%>%
  arrange(-DMIN)%>%
  select(ends_with("UDE"),contains("TEMP"),YEAR,COMM,TOTWGT)%>%
  rename(species=COMM)%>%
  head ## ORDER MATTERS
```

##filter --------

select specific rows with the *filter* function

```{r}
filter(rvdata,YEAR==2010)%>%head
```

```{r}
filter(rvdata,YEAR==2010)%>%
  arrange(COMM,-TOTWGT)%>%
  glimpse
```

```{r}
filter(rvdata,YEAR==2010,XTYPE==1,COMM %in% c("COD(ATLANTIC)","AMERICAN LOBSTER"))%>%
  arrange(COMM,-TOTWGT)%>%
  glimpse
```

```{r}
filter(rvdata,YEAR==2010,XTYPE==1,COMM %in% c("COD(ATLANTIC)","AMERICAN LOBSTER"))%>%
  arrange(COMM,-TOTWGT)%>%
  select(COMM,YEAR)%>%
  glimpse
```

```{r}
filter(rvdata,between(YEAR,2010,2012),between(TOTWGT,10,14),COMM=="AMERICAN LOBSTER")
```

##sample -------

You can also use the *sample* function to select a random subset or fraction of code

sample five rows
```{r}
rvdata%>%sample_n(5)
```

sample 2% of the data
```{r}
rvdata%>%sample_frac(0.02)%>%head
```

##mutate -----

create a new column with the same dimensions as you started

```{r}
rvdata%>%mutate(dAVE = (DMIN+DMAX)/2)%>%head
```

```{r}
rvdata%>%mutate(dAVE = mean(c(DMIN,DMAX)))%>%head 
```

```{r}
rvdata%>%mutate(dAVE = rowMeans(select(.,starts_with("DM"))))%>%head
```

```{r}
rvdata%>%mutate(dAVE = (DMIN+DMAX)/2,
                wAVE = TOTWGT/TOTNO)%>%head
```

##summarise ------------

*summarise* data. In this case whatever is piped is manipuated to a single value. 
```{r}
rvdata%>%summarise(mn=mean(TOTWGT,na.rm=T),
                   sd=sd(TOTWGT,na.rm=T))
```

##group_by -------

*group_by* is a very useful function that allows you to aggregate the data inline before any selections, mutations or summaries are extracted. 

for example lets revisit sample. In this case I want to sample 5 observations per year and per species. In this case I use sample with replacement because some species are observed fewer than 5 times in a given year. 

```{r}
rvdata%>%group_by(COMM,YEAR)%>%sample_n(5,replace=T)%>%ungroup()%>%glimpse
```

average weight for Lobster and Cusk per year using *summarise*

```{r}
rvdata%>%
  filter(COMM %in% c("AMERICAN LOBSTER","CUSK"))%>%
  group_by(COMM,YEAR)%>%
  summarise(mn=mean(TOTWGT,na.rm=T),
            sd=sd(TOTWGT,na.rm=T))%>%glimpse
```

create a variable that simulates distance and then create a standardized tow weight based on the simulated to distance for lobster and cusk and get the summary per year for 2013:2016

```{r}
rvdata%>%
  filter(COMM %in% c("AMERICAN LOBSTER","CUSK"),YEAR %in% 2013:2016)%>%
  mutate(DIST = sample(x=seq(from= 1,to = 3,by=0.01),size=n(),replace=T),
                STDWGT = TOTWGT*1.75/DIST)%>%
  group_by(COMM,YEAR)%>%
  summarise(mn=mean(STDWGT))
```

## Example 

Lets try to calculate species richness based on the RV survey. The data has each species as a row so each set has multiple rows. We can also scale the diversity within each year to a 0-1 scale. This requires that data is grouped by set and year then ungrouped and regrouped again by year so the mutate can be applied coorectly (i.e., max for a given year is used in the scaling). We also need to bring the coordinates back into this calculation but can't bring them through the grouping so we merge back in after the calculations are finished

```{r}
diversity <- rvdata%>%
  filter(XTYPE==1,grepl("NED",.$MISSION))%>%
  group_by(YEAR,MISSION,SETNO)%>%
  summarise(abundance = sum(TOTNO),
            div = n())%>% # n() returns the length of the data. 
  ungroup()%>% 
  group_by(YEAR)%>% #regroup
  mutate(pdiv=div/max(div,na.rm=T))%>%
  ungroup()%>%
  right_join(.,rvdata%>%distinct(MISSION,SETNO,.keep_all=T)%>%
               select(ends_with("UDE"),MISSION,SETNO,YEAR),
             by=c("MISSION","SETNO","YEAR"))%>% #merge back in the coordinates 
  rename(long=LONGITUDE,lat=LATITUDE,year=YEAR)%>%
  filter(!is.na(div))%>%
  data.frame()
```

Lets see the results on a map
```{r}
makemap(diversity%>%filter(between(year,2015,2018),!is.na(pdiv))%>%dplyr::select(long,lat,pdiv))

```

