---
title: "Making maps with Base R and the oce package"
output: html_notebook
---

From the RV data analysis, a common goal (for a report or paper) would be to summarize in a plot such as a map. This is easy to do using base R (where "base R" refers to using "base R *graphics*", i.e. *not* `ggplot2`), especially with the help of some of the mapping functionality in the `oce` package.

## Getting the data

First, we run the script that Ryan wrote to get the data in our workspace:

```{r}
source('DataSetup.R')
d <- plotdata    # because I'm lazy
```

We can see what's in the data:
```{r}
str(d)
```

The simplest "map" is acheived by just plotting the lon/lat data as x/y:

```{r}
with(d, plot(long, lat))
```
However, this isn't a very useful map, because it contains not spatial context, other than the values of long and lat. Things like coasts and bathymetry would make this make a lot more informative.

## Mapping with oce

Much of the mapping capabilities in `oce` are built around `coastline` objects. These are simple objects that define the lon/lat of the points that trace out a landmass coastline. For ease, there are several global coastline data sets included with `oce` (and its sister package, `ocedata`), which are: `coastlineWorld` (lowest resolution, in `oce`), `coastlineWorldMedium` ("medium" resolution, contained in `ocedata`), and `coastlineWorldFine` (highest res, in `ocedata`, suitable for limited domains).

We can get the various coastlines using the `data()` function, which loads built-in data sets:
```{r}
library(oce)
library(ocedata)
data(coastlineWorld)
data(coastlineWorldMedium)
data(coastlineWorldFine)
```

Thanks to the OO (object orientation), there exists a generic `plot()` method for coastline objects:
```{r}
plot(coastlineWorld)
```

Using a basic "long/lat" or something resembling a "Mercator" projection (i.e no real scaling on the lon/lat axes) makes the world plot look wonky because of distortion at high latitudes.

For a global plot, it is almost always better to use *projected* coordinates, which can be acheived by passing the `projection=` argument in `plot.coastline()`:
```{r}
plot(coastlineWorld, projection='+proj=moll', col='grey') # use a Molleweide projection
```

This is similar to using the specific function, `mapPlot()` which handles the projection directly. In most of the following I will use `mapPlot()` instead of the `plot,coastline-method` generic function, since it is clearer.

The syntax for the projection, i.e .the `+proj=` is from the PROJ4 library, which is included in the package from the GDAL libraries (and `rgdal` package). Read the help page for `?mapPlot()` for info and links to more resources.

Some other nice projections:
```{r}
mapPlot(coastlineWorld, projection='+proj=wintri', col='grey')
```

```{r}
mapPlot(coastlineWorld, projection='+proj=vandg', col='grey')
```

```{r}
mapPlot(coastlineWorld, projection='+proj=robin', col='grey')
```

```{r}
mapPlot(coastlineWorld, projection='+proj=igh', col='grey')
```

What about the Arctic?
```{r}
mapPlot(coastlineWorld, projection='+proj=stere +lat_0=90', longitudelim = c(-180, 180), latitudelim = c(60, 90), col='grey')
```


## Non-global maps

For most of us, world maps are not the end goal for when we want to plot data (e.g. the RV survery). Here I'll go over some of the arguments to use for `plot,coastline-method` and `mapPlot()` to make nice Atlantic maps.

You can compare the `coastlineWorld` and `coastlineWorldFine` data sets to see what different it makes for a Maritimes view:
```{r}
par(mfrow=c(1, 3))
plot(coastlineWorld, clatitude=45, clongitude=-62, span=2000, col='grey')
plot(coastlineWorldMedium, clatitude=45, clongitude=-62, span=2000, col='grey')
plot(coastlineWorldFine, clatitude=45, clongitude=-62, span=2000, col='grey')
```

Note the `clatitude/clongitude` and `span` arguments for `plot,coastline-method`. 

Let's make a nice map that we can overlay the RV data on. This time we'll use `mapPlot` but just with the 

```{r}
mapPlot(coastlineWorldFine, longitudelim = c(-66, -58), latitudelim=c(40, 50),
        projection='+proj=lcc +lon_0=-62', grid=FALSE, col='grey')
m <- recordPlot()
```

We can add graticules (a grid), using the `mapGrid()` function, and then we can add points, using the "helper" function `mapPoints()`:
```{r}
mapPlot(coastlineWorldFine, longitudelim = c(-66, -58), latitudelim=c(40, 50),
        projection='+proj=lcc +lon_0=-62', grid=FALSE, col='grey')
mapGrid(2, 2)
with(d, mapPoints(long, lat))
```

And what if we want to add detail (species, colours)? It's just like building up a normal base R plot:
```{r}
mapPlot(coastlineWorldFine, longitudelim = c(-66, -58), latitudelim=c(40, 50),
        projection='+proj=lcc +lon_0=-62', grid=FALSE, col='grey')
mapGrid(2, 2)
with(d[d$species == 'AMERICAN LOBSTER',], mapPoints(long, lat, pch=19, cex=0.5, col=1))
with(d[d$species == 'COD(ATLANTIC)',], mapPoints(long, lat, pch=19, cex=0.5, col=2))
with(d[d$species == 'CUSK',], mapPoints(long, lat, pch=19, cex=0.5, col=3))
```

